---
title: "Comparing metabolomic and transcriptomic data."
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
output:
  html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: tango
    keep_md: false
    mode: selfcontained
    number_sections: true
    self_contained: true
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  rmdformats::readthedown:
    code_download: true
    code_folding: show
    df_print: paged
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: tango
    width: 300
    keep_md: false
    mode: selfcontained
    toc_float: true
  BiocStyle::html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: tango
    keep_md: false
    mode: selfcontained
    toc_float: true
---

<style type="text/css">
body, td {
  font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
 font-size: 16px
}
</style>

```{r options, include=FALSE}
library("hpgltools")
tt <- devtools::load_all("~/hpgltools")
knitr::opts_knit$set(width=120,
                     progress=TRUE,
                     verbose=TRUE,
                     echo=TRUE)
knitr::opts_chunk$set(error=TRUE,
                      dpi=96)
old_options <- options(digits=4,
                       stringsAsFactors=FALSE,
                       knitr.duplicate.label="allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size=10))
rundate <- format(Sys.Date(), format="%Y%m%d")
previous_file <- ""
ver <- format(Sys.Date(), "%Y%m%d")

##tmp <- sm(loadme(filename=paste0(gsub(pattern="\\.Rmd", replace="", x=previous_file), "-v", ver, ".rda.xz")))
rmd_file <- "index.Rmd"
```

# Comparing metabolomic and transcriptomic data

One of the recent discussions surrounding the TMRC3 data is centered
on a series of metabolomic experiments.  I have not yet spent the time
to truly understand the experiment performed, I am hoping to use the
process of writing this document to amend that.  In addition, this
document will hopefully provide some of the pieces of inquiry useful
for informing the overall discussion.

## The metabolomic data

This portion of the data is coming from Alejandro's paper:
"Pharmacometabolomics of Meglumine Antimoniate in Patients
With Cutaneous Leishmaniasis."  I therefore am dropping a copy of this
paper in the 'downloaded_data' directory and reading it before I make
any more assumptions.

# Collecting the transcriptomic results

There is one super-important caveat when collecting the transcriptomic
data: the TMRC3 data includes two different drug regimes with
apparently two very different effects on the person's metabolism.  As
a result, I cannot use the set of all samples, but must limit it to
the samples which used glucantine and _not_ miltefosine.

In our most recent conversation, the focus shifted to two primary
contrasts performed in the data:

1.  The comparison of visit 3 / visit 1 in monocytes treated with glucantine.
2.  The comparison of visit 3 / visit 1 in biopsies treated with glucantine.

I therefore copied the DE table data structures to the rda/ directory
and will do some simple tests on it.

### Collecting monocytes

```{r collecting_data_monocytes}
monocyte <- new.env()
load(file = "rda/visit_monocyte_gluc_table-v202011.rda", envir = monocyte)
monocyte <- monocyte[["visit_monocyte_gluc_table"]]

monocyte_topn <- extract_significant_genes(monocyte, n = 100, according_to = "deseq")
monocyte_v2v1_up <- monocyte_topn[["deseq"]][["ups"]][["v2v1"]]
monocyte_v2v1_down <- monocyte_topn[["deseq"]][["downs"]][["v2v1"]]
monocyte_v3v1_up <- monocyte_topn[["deseq"]][["ups"]][["v3v1"]]
monocyte_v3v1_down <- monocyte_topn[["deseq"]][["downs"]][["v3v1"]]
```

### Collecting data biopsies

```{r collecting_data_biopsies}
biopsy <- new.env()
load(file = "rda/visit_biopsy_gluc_table-v202011.rda", envir = biopsy)
biopsy <- biopsy[["visit_biopsy_gluc_table"]]

biopsy_topn <- extract_significant_genes(biopsy, according_to = "deseq")
biopsy_v2v1_up <- biopsy_topn[["deseq"]][["ups"]][["v2v1"]]
dim(biopsy_v2v1_up)
biopsy_v2v1_down <- biopsy_topn[["deseq"]][["downs"]][["v2v1"]]
dim(biopsy_v2v1_down)

biopsy_v3v1_up <- biopsy_topn[["deseq"]][["ups"]][["v3v1"]]
dim(biopsy_v3v1_up)
biopsy_v3v1_down <- biopsy_topn[["deseq"]][["downs"]][["v3v1"]]
dim(biopsy_v3v1_down)
```

## GSEA of visits

### Monocytes

```{r monocyte_visit_gsea}
monocyte_v2v1_gpup <- simple_gprofiler(monocyte_v2v1_up)
monocyte_v2v1_gpup[["pvalue_plots"]][["bpp_plot_over"]]
monocyte_v2v1_gpup[["pvalue_plots"]][["mfp_plot_over"]]
monocyte_v2v1_gpup[["pvalue_plots"]][["kegg_plot_over"]]

monocyte_v2v1_gpdown <- simple_gprofiler(monocyte_v2v1_down)
monocyte_v2v1_gpdown[["pvalue_plots"]][["bpp_plot_over"]]
monocyte_v2v1_gpdown[["pvalue_plots"]][["mfp_plot_over"]]
monocyte_v2v1_gpdown[["pvalue_plots"]][["kegg_plot_over"]]


monocyte_v3v1_gpup <- simple_gprofiler(monocyte_v3v1_up)
monocyte_v3v1_gpup[["pvalue_plots"]][["bpp_plot_over"]]
monocyte_v3v1_gpup[["pvalue_plots"]][["mfp_plot_over"]]
monocyte_v3v1_gpup[["pvalue_plots"]][["kegg_plot_over"]]

monocyte_v3v1_gpdown <- simple_gprofiler(monocyte_v3v1_down)
monocyte_v3v1_gpdown[["pvalue_plots"]][["bpp_plot_over"]]
monocyte_v3v1_gpdown[["pvalue_plots"]][["mfp_plot_over"]]
monocyte_v3v1_gpdown[["pvalue_plots"]][["kegg_plot_over"]]
```


### Biopsys

```{r biopsy_visit_gsea}
biopsy_v2v1_gpup <- simple_gprofiler(biopsy_v2v1_up)
biopsy_v2v1_gpup[["pvalue_plots"]][["bpp_plot_over"]]
biopsy_v2v1_gpup[["pvalue_plots"]][["mfp_plot_over"]]
biopsy_v2v1_gpup[["pvalue_plots"]][["kegg_plot_over"]]

biopsy_v2v1_gpdown <- simple_gprofiler(biopsy_v2v1_down)
biopsy_v2v1_gpdown[["pvalue_plots"]][["bpp_plot_over"]]
biopsy_v2v1_gpdown[["pvalue_plots"]][["mfp_plot_over"]]
biopsy_v2v1_gpdown[["pvalue_plots"]][["kegg_plot_over"]]


biopsy_v3v1_gpup <- simple_gprofiler(
    biopsy_v3v1_up,
    excel = glue::glue("excel/v3v1_up_gprofiler-v{ver}.xlsx"))
biopsy_v3v1_gpup[["pvalue_plots"]][["bpp_plot_over"]]
biopsy_v3v1_gpup[["pvalue_plots"]][["mfp_plot_over"]]
biopsy_v3v1_gpup[["pvalue_plots"]][["kegg_plot_over"]]
biopsy_v3v1_gpup[["pvalue_plots"]][["reactome_plot_over"]]

biopsy_v3v1_gpdown <- simple_gprofiler(
    biopsy_v3v1_down,
    excel = glue::glue("excel/v3v1_down_gprofiler-v{ver}.xlsx"))
biopsy_v3v1_gpdown[["pvalue_plots"]][["bpp_plot_over"]]
biopsy_v3v1_gpdown[["pvalue_plots"]][["mfp_plot_over"]]
biopsy_v3v1_gpdown[["pvalue_plots"]][["kegg_plot_over"]]
```

# GSVA of the upstream data sets

## Monocytes

## Biopsies

# KEGG Mapping and Cross Referencing

In order to color the various KEGG path diagrams, we will need to make
some bridges across the various databases, though I think the bridges
will need to be less interesting than they were when we thought it was
L. panamensis.

The things we will need (I think):

1.  Mapping from the Ensembl IDs to NCBI IDs (load_biomart), and the KEGG
IDs to NCBI (load_kegg).
2.  Table of kegg and gene IDs.
3.  Set of pathways of interest.

## Starter set

Lets make it easy on ourselves and just color some KEGG pictures with
the observed genes from the biopsy up/down.

```{r kegg_testing}
updown_genes <- rbind(biopsy_v3v1_up, biopsy_v3v1_down)

wanted_columns <- c("ensembl_gene_id", "ensembl_gene_id_version",
                    "entrez_trans_name", "entrezgene_description",
                    "entrezgene_accession", "entrezgene_id")
hs_annot <- load_biomart_annotations(archive = FALSE, gene_requests = wanted_columns,
                                     include_lengths = FALSE, overwrite = TRUE)[["annotation"]]
kegg_annot <- load_kegg_annotations(abbreviation = "hsa")

## hs_annot says that hsa:7105 is tetraspannin.6, how about kegg?
test_id <- "7105"
test_hsa_idx <- hs_annot[["entrezgene_id"]] == test_id
head(hs_annot[test_hsa_idx, ], n = 1)

test_kegg_idx <- kegg_annot[["kegg_geneid"]] == paste0("hsa:", test_id)
kegg_annot[test_kegg_idx, ]


interesting_stuff <- merge(updown_genes, hs_annot,
                           by.x = "ensemblgeneid", by.y = "ensembl_gene_id", all.x = TRUE)
dim(updown_genes)
dim(interesting_stuff)
## So, there appear to be a few duplicates, that should be ok.
interesting_stuff[["kegg_prefix_id"]] <- paste0("hsa:", interesting_stuff[["entrezgene_id"]])
## interesting_stuff <- merge(interesting_stuff, kegg_annot, by.x = "temp_id", by.y = "kegg_geneid")
## Hey doofus, kegg uses 'hsa:', followed by the NCBI entrez gene ID,
## I don't need to cross reference the darn kegg IDs at all.
## well, ok, lets see if we get any fun colored paths...

pathview_input <- interesting_stuff
testing <- simple_pathview(gene_input = pathview_input, species = "hsa",
                           fc_column = "deseq_logfc",
                           id_column = "entrezgene_id")
                           ## id_column = "kegg_prefix_id")
```

# Add compounds and genes

We have a csv/tsv file of pathways and compounds.  It is in a peculiar
format and does not have any information about the range of
increase/decrease.  As a result, I do not think we can meaningfully
scale the data; but we can extract all the compound IDs and set their color.

```{r get_compounds}
compound_start <- readr::read_csv("downloaded_data/joint_metaboanalyst.csv")
pile_of_compounds_and_junk <- compound_start[[2]]
split_stuff <- strsplit(x = pile_of_compounds_and_junk, split = "; ")
compounds <- c()
for (i in 1:length(split_stuff)) {
  internal <- split_stuff[[i]]
  cpd_idx <- grep("^cpd", internal)
  internal <- internal[cpd_idx]
  internal_ids <- stringi::stri_replace_all_fixed(internal, ";", "")
  internal_ids <- gsub(pattern = "^cpd:", replacement = "", x = internal_ids)
  compounds <- c(internal_ids, compounds)
}
compounds <- unique(compounds)
compound_input <- rep(10, length(compounds))
names(compound_input) <- compounds


testing <- simple_pathview(gene_input = pathview_input, compound_input = compound_input,
                           species = "hsa",
                           fc_column = "deseq_logfc",
                           id_column = "entrezgene_id")
```


```{r saveme, eval=FALSE}
pander::pander(sessionInfo())
message(paste0("This is hpgltools commit: ", get_git_commit()))
this_save <- paste0(gsub(pattern="\\.Rmd", replace="", x=rmd_file), "-v", ver, ".rda.xz")
message(paste0("Saving to ", this_save))
tmp <- sm(saveme(filename=this_save))
```
