---
title: "Comparing metabolomic and transcriptomic data."
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
output:
  html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: tango
    keep_md: false
    mode: selfcontained
    number_sections: true
    self_contained: true
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  rmdformats::readthedown:
    code_download: true
    code_folding: show
    df_print: paged
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: tango
    width: 300
    keep_md: false
    mode: selfcontained
    toc_float: true
  BiocStyle::html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: tango
    keep_md: false
    mode: selfcontained
    toc_float: true
---

<style type="text/css">
body, td {
  font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
 font-size: 16px
}
</style>

```{r options, include=FALSE}
library("hpgltools")
tt <- devtools::load_all("~/hpgltools")
knitr::opts_knit$set(width=120,
                     progress=TRUE,
                     verbose=TRUE,
                     echo=TRUE)
knitr::opts_chunk$set(error=TRUE,
                      dpi=96)
old_options <- options(digits=4,
                       stringsAsFactors=FALSE,
                       knitr.duplicate.label="allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size=10))
rundate <- format(Sys.Date(), format="%Y%m%d")
previous_file <- ""
ver <- format(Sys.Date(), "%Y%m%d")

##tmp <- sm(loadme(filename=paste0(gsub(pattern="\\.Rmd", replace="", x=previous_file), "-v", ver, ".rda.xz")))
rmd_file <- "index.Rmd"
```

# Comparing metabolomic and transcriptomic data

One of the recent discussions surrounding the TMRC3 data is centered
on a series of metabolomic experiments.  I have not yet spent the time
to truly understand the experiment performed, I am hoping to use the
process of writing this document to amend that.  In addition, this
document will hopefully provide some of the pieces of inquiry useful
for informing the overall discussion.

## The metabolomic data

This portion of the data is coming from Alejandro's paper:
"Pharmacometabolomics of Meglumine Antimoniate in Patients
With Cutaneous Leishmaniasis."  I therefore am dropping a copy of this
paper in the 'downloaded_data' directory and reading it before I make
any more assumptions.

# Collecting the transcriptomic results

There is one super-important caveat when collecting the transcriptomic
data: the TMRC3 data includes two different drug regimes with
apparently two very different effects on the person's metabolism.  As
a result, I cannot use the set of all samples, but must limit it to
the samples which used glucantine and _not_ miltefosine.

In our most recent conversation, the focus shifted to two primary
contrasts performed in the data:

1.  The comparison of visit 3 / visit 1 in monocytes treated with glucantine.
2.  The comparison of visit 3 / visit 1 in biopsies treated with glucantine.

I therefore copied the DE table data structures to the rda/ directory
and will do some simple tests on it.

```{r collecting_data}
monocyte <- new.env()
load(file = "rda/visit_monocyte_gluc_table-v202011.rda", envir = monocyte)
monocyte <- monocyte[["visit_monocyte_gluc_table"]]
load("rda/visit_biopsy_gluc_table-v202011.rda")

monocyte_topn <- extract_significant_genes(monocyte, n = 100, according_to = "deseq")
monocyte_v2v1_up <- monocyte_topn[["deseq"]][["ups"]][["v2v1"]]
monocyte_v2v1_down <- monocyte_topn[["deseq"]][["downs"]][["v2v1"]]
monocyte_v3v1_up <- monocyte_topn[["deseq"]][["ups"]][["v3v1"]]
monocyte_v3v1_down <- monocyte_topn[["deseq"]][["downs"]][["v3v1"]]
```

## GSEA of visits

```{r monocyte_visit_gsea}
monocyte_v2v1_gpup <- simple_gprofiler(monocyte_v2v1_up)
monocyte_v2v1_gpup[["pvalue_plots"]][["bpp_plot_over"]]
monocyte_v2v1_gpup[["pvalue_plots"]][["mfp_plot_over"]]
monocyte_v2v1_gpup[["pvalue_plots"]][["kegg_plot_over"]]

monocyte_v2v1_gpdown <- simple_gprofiler(monocyte_v2v1_down)
monocyte_v2v1_gpdown[["pvalue_plots"]][["bpp_plot_over"]]
monocyte_v2v1_gpdown[["pvalue_plots"]][["mfp_plot_over"]]
monocyte_v2v1_gpdown[["pvalue_plots"]][["kegg_plot_over"]]


monocyte_v3v1_gpup <- simple_gprofiler(monocyte_v3v1_up)
monocyte_v3v1_gpup[["pvalue_plots"]][["bpp_plot_over"]]
monocyte_v3v1_gpup[["pvalue_plots"]][["mfp_plot_over"]]
monocyte_v3v1_gpup[["pvalue_plots"]][["kegg_plot_over"]]

monocyte_v3v1_gpdown <- simple_gprofiler(monocyte_v3v1_down)
monocyte_v3v1_gpdown[["pvalue_plots"]][["bpp_plot_over"]]
monocyte_v3v1_gpdown[["pvalue_plots"]][["mfp_plot_over"]]
monocyte_v3v1_gpdown[["pvalue_plots"]][["kegg_plot_over"]]
```




```{r saveme, eval=FALSE}
pander::pander(sessionInfo())
message(paste0("This is hpgltools commit: ", get_git_commit()))
this_save <- paste0(gsub(pattern="\\.Rmd", replace="", x=rmd_file), "-v", ver, ".rda.xz")
message(paste0("Saving to ", this_save))
tmp <- sm(saveme(filename=this_save))
```
